Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-15 linear_iso.tex
about to edit numerical

\section{Analytical discussion of vertically isothermal
  disks in the framework of the vertically global shearing
  box}\label{analytical}  
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant in the radially-local
approximation. % We examine isothermal perturbations ($\beta =
% 0$) before considering the effect of finite thermal relaxation. 
In order to facilitate a fully analytic discussion, we first introduce a
number of approximations below that allows a simple governing equation
to be derived and compared with previous works.  However, all of the 
approximations described in
\S\ref{vgsb_approx}---\S\ref{analytic_relax} are relaxed in our
numerical study.

\subsection{Relation to the vertically global
  shearing box}\label{vgsb_approx} 
We connect the linearized equations 
Eq. \ref{lin_mass}---\ref{lin_energy} to the vertically global  
shearing box (VGSB) formalism developed by \cite{mcnally14}. The VGSB
is an extension of the standard shearing box \citep{goldreich65} to
allow the shear flow in the shearing box to be height-dependent. 
The difference between the radially local
approximation and the VGSB is that the equilibrium disk
structure is assumed to be homogeneous in the radial ($x$) direction,
as in the 
standard shearing box.

@@

After setting $\hat{g}_c=0$ in Eq. \ref{lin_mass}, \ref{lin_vx} and
\ref{lin_energy}, we can eliminate variables % eliminate the perturbed velocities between
% the linearized equations 
to obtain a pair or ordinary differential
equations for $W,Q$ as
\begin{align}
  &\frac{\sigma^2}{c_s^2}Q + \frac{\sigma^2k_x^2}{D}W \notag\\ 
  &=\left(\frac{\ii
      k_x r}{D}\frac{d\Omega^2}{dz} -
    \frac{d\ln{\rho}}{dz}\right)\left[\frac{dW}{dz} +
    \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right] \notag\\
  &- \frac{d^2W}{dz^2} - \frac{d\ln{\rho}}{dz}\left(\frac{dW}{dz} -
    \frac{dQ}{dz}\right) - \frac{d^2\ln{\rho}}{dz^2}\left(W-Q\right),\label{ode_w}\\
  &\sigma^2W - \frac{\gamma}{\Gamma}\sigma^2 Q +
  \frac{\ii\sigma}{t_c}\left(W-\frac{Q}{\Gamma}\right)\notag\\
  &=c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma} - 1\right) 
  \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right],\label{ode_Q} 
\end{align}
where
\begin{align}
  D \equiv \kappa^2 - \sigma^2,
\end{align} 
and we have replaced the notation $\p_z$ by $d/dz$ since we are
considering solutions at a fixed radius.

Extrapolating results obtained from the VGSB to the global disk has
limitations. This is because within the VGSB framework, $d\Omega/dz$
can be freely specified. %  In effect, by adopting the VGSB framework, we are studying
% dynamics in the standard shearing box subject to an additional
% azimuthal force $\delta v_zr d \Omega/dz$ with $d\Omega/dz$ as an
% input parameter. 
%For consistency with the global disk, 
We choose $d\Omega/dz$ to be that due to the thermal structure in 
the global disk (Eq. \ref{vertical_shear}). However, background radial
gradient terms are ignored elsewhere in the VGSB (which is done by
setting $\hat{g}_c=0$).

%the above inconsistency

We further discuss caveats of the VGSB in Appendix
\ref{global_corr}. We find that for perturbations with small radial
lengthscales ($k_xH\gg 1$) and fast thermal relaxation ($\beta\ll1$),
which characterizes the VSI, the VGSB framework is adequate for the 
purpose of a simple analytical treatment.

\subsection{Low-frequency and nearly-Keplerian
  approximation}\label{approx_gov} 
We now derive a governing equation with further approximations:
\begin{enumerate}
\item We consider eigenvalues such that
  $|\sigma^2|\ll \kappa^2$. Then we can replace $D=\kappa^2 -\sigma^2\to
  \kappa^2$. % The problem simplifies because the eigenvalue now only
  % appears in the last term in Eq. \ref{iso_ode}. %stoll kley see low
  % freq 
  We call this the low-frequency approximation. 
\item We are mainly interested in the effect of vertical shear, i.e. $q\neq
  0$. This is already captured explicitly in the linearized equations through
  the term $d\Omega/dz$. We therefore simplify 
  further by ignoring the vertical dependence of $\Omega$ and $\kappa$
  elsewhere. For clarity, we replace $\kappa^2$ with
  $\Omega_k^2$. We call this the nearly-Keplerian approximation.

@@

It is convenient to introduce the dimensionless variables
\begin{align}
  \hat{z} = z/H,\quad \hat{k}=k_xH, \quad \hat{\sigma} =\hat{\omega} +
  \ii\hat{\nu}= \sigma/\Omega_k,
\end{align}
where $\hat{\omega}$ and $\hat{\nu}$ are real. 
Eq. \ref{vertiso_gov} becomes 
\begin{align}
  0=& \delta v_z ^{\prime\prime} + \left(1 + \ii\epsilon
    q\hat{k}\right)\ln\rho^{\prime}\delta v_z^\prime \notag\\
  &+
  \left[\hat{\sigma}^2\left(\hat{k}^2+\chi\right) +
    \left(  \chi + \ii \epsilon q\hat{k}\right)\ln\rho^{\prime\prime}\right.
  \notag\\
  &\left.- \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii\epsilon
      q\hat{k}\right)\left(1 - \chi\right)\right]\delta v_z,\label{vertiso_gov_nondim}
\end{align}
where $^\prime$ denotes $d/d\zhat$.%  Eq. \ref{vertiso_gov_nondim} will
% be the basis for our discussion of vertically isothermal disks.

In deriving Eq. \ref{vertiso_gov}, the vertical dependence of
$D=\kappa^2(z)-\sigma^2$ was ignored because the replacement
$D\to\Omega_k^2$ was made before eliminating variables in favor of
$\delta v_z$. In Appendix \ref{adia_improve} we show that making this
replacement after eliminating variables, which involves $dD/dz$,
introduces terms $O(\epsilon^2)$. Since we are interested in 
thin disks ($\epsilon\ll 1 $), e.g. protoplaneary disks where 
$\epsilon\lesssim 0.1$ \citep{chiang10}, these terms introduce
unnecessary complexity for the present discussion. % However, this
% approximation will be relaxed in our numerical study.

% Note that we can generalize Eq. \ref{vertiso_gov_nondim} non-Keplerian
% rotation profiles by redefining $\khat$ as $k_xc_s/\kappa$,
% $\hat{\sigma}$ as $\sigma/\kappa$, and  $\epsilon$ as $c_s/r\kappa$.

% This expression can be simplified in the limit $|q|\ll 1$, 
%  \begin{align}\label{simple_growth}
%    \frac{\imag(\sigma)}{\Omega_k} = \pm \sqrt{M} 
%    \frac{q\epsilon\hat{k}}{2\sqrt{1+\hat{k}^2}} \quad\quad (|q|\ll 1), 
%  \end{align}
%  with $M$ being an integer.

% In Appendix \ref{pert_theory} we give an alternative method to
% infer instability in the presence of vertical shear without explicitly
% solving the governing equation.

>>>>

\section{Analytical discussion of vertically isothermal
  disks in the framework of the vertically global shearing
  box}\label{analytical}  
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant in the radially-local
approximation. 
In order to facilitate a fully analytic discussion, we first introduce a
number of approximations below that allows a simple governing equation
to be derived and compared with previous works.  However, all of the 
approximations described below are relaxed in our
numerical study.

\subsection{Relation to the vertically global
  shearing box}\label{vgsb_approx} 
We connect the linearized equations 
Eq. \ref{lin_mass}---\ref{lin_energy} to the vertically global  
shearing box (VGSB) formalism developed by \cite{mcnally14}. The VGSB
is an extension of the standard shearing box \citep{goldreich65} to 
background shear flows that are height dependent. 
As in the  standard shearing box, in the VGSB the equilibrium disk
structure is assumed to be homogeneous in the radial ($x$) direction.

@@

After setting $\hat{g}_c=0$ in Eq. \ref{lin_mass}, \ref{lin_vx} and
\ref{lin_energy}, we can eliminate variables 
to obtain a pair or ordinary differential
equations for $W,Q$ as
\begin{align}
  &\frac{\sigma^2}{c_s^2}Q + \frac{\sigma^2k_x^2}{D}W \notag\\ 
  &=\left(\frac{\ii
      k_x r}{D}\frac{d\Omega^2}{dz} -
    \frac{d\ln{\rho}}{dz}\right)\left[\frac{dW}{dz} +
    \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right] \notag\\
  &- \frac{d^2W}{dz^2} - \frac{d\ln{\rho}}{dz}\left(\frac{dW}{dz} -
    \frac{dQ}{dz}\right) - \frac{d^2\ln{\rho}}{dz^2}\left(W-Q\right),\label{ode_w}\\
  &\sigma^2W - \frac{\gamma}{\Gamma}\sigma^2 Q +
  \frac{\ii\sigma}{t_c}\left(W-\frac{Q}{\Gamma}\right)\notag\\
  &=c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma} - 1\right) 
  \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right],\label{ode_Q} 
\end{align}
where
\begin{align}
  D \equiv \kappa^2 - \sigma^2,
\end{align} 
and we have replaced the notation $\p_z$ by $d/dz$ since we are
considering solutions at a fixed radius.

Extrapolating results obtained from the VGSB to the global disk has
limitations. This is because within the VGSB framework, $d\Omega/dz$
can be freely specified. 
We choose $d\Omega/dz$ to be that due to the thermal structure in 
the global disk (Eq. \ref{vertical_shear}). However, background radial
gradient terms are ignored elsewhere in the VGSB (which is done by
setting $\hat{g}_c=0$).

We further discuss caveats of the VGSB in Appendix
\ref{global_corr}. We find that for fast thermal relaxation
($\beta\ll1$) and perturbations with small radial
lengthscales ($k_xH\gg 1$), which characterizes the VSI, the VGSB
framework is adequate for the purpose of a simple analytical
treatment.

\subsection{Low-frequency and nearly-Keplerian
  approximation}\label{approx_gov} 
We now derive a governing equation with further approximations:
\begin{enumerate}
\item We consider eigenvalues such that
  $|\sigma^2|\ll \kappa^2$. Then we can replace $D=\kappa^2 -\sigma^2\to
  \kappa^2$. %\citep[e.g.][]{lin10}. 
  We call this the low frequency approximation. %physically inertial
                                %waves 
\item We are mainly interested in the effect of vertical shear, i.e. $q\neq
  0$. This is already captured explicitly in the linearized equations through
  the term $d\Omega/dz$. We therefore simplify 
  further by ignoring the vertical dependence of $\Omega$ and $\kappa$
  elsewhere. For clarity, we replace $\kappa^2$ with
  $\Omega_k^2$. We call this the nearly-Keplerian approximation.

@@

It is convenient to introduce the dimensionless variables
\begin{align}
  \hat{z} = z/H,\quad \hat{k}=k_xH, \quad \hat{\sigma} =\hat{\omega} +
  \ii\hat{\nu}= \sigma/\Omega_k,
\end{align}
where $\hat{\omega}$ and $\hat{\nu}$ are real. 
Eq. \ref{vertiso_gov} becomes 
\begin{align}
  0=& \delta v_z ^{\prime\prime} + \left(1 + \ii\epsilon
    q\hat{k}\right)\ln\rho^{\prime}\delta v_z^\prime \notag\\
  &+
  \left[\hat{\sigma}^2\left(\hat{k}^2+\chi\right) +
    \left(  \chi + \ii \epsilon q\hat{k}\right)\ln\rho^{\prime\prime}\right.
  \notag\\
  &\left.- \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii\epsilon
      q\hat{k}\right)\left(1 - \chi\right)\right]\delta v_z,\label{vertiso_gov_nondim}
\end{align}
where $^\prime$ denotes $d/d\zhat$.

In deriving Eq. \ref{vertiso_gov}, the vertical dependence of
$D=\kappa^2(z)-\sigma^2$ was ignored because the replacement
$D\to\Omega_k^2$ was made before eliminating variables in favor of
$\delta v_z$. In Appendix \ref{adia_improve} we show that making this
replacement after eliminating variables, which involves $dD/dz$,
introduces terms $O(\epsilon^2)$. Since we are interested in 
thin disks ($\epsilon\ll 1 $), e.g. protoplaneary disks where 
$\epsilon\lesssim 0.1$ \citep{chiang10}, these terms introduce
unnecessary complexity for the present discussion.

### 2015-03-15 numerical.tex
about to edit numerical

\subsection{VSI with rapid thermal relaxation}\label{vertiso_pertiso} 
We first calculate the VSI in a disk with rapid thermal relaxation by
setting $\beta=10^{-3}$. (Smaller values yield the similar results.) This
is similar to strictly isothermal perturbations in a vertically
isothermal disk considered in \cite{nelson13} and \cite{mcnally14}.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_modes_iso_kx30_analytic.ps}
  \caption{Same as Fig. \ref{compare_modes_iso_kx10} but for $\khat=30$
    \label{compare_modes_iso_kx30}. Examples of surface modes are
    marked by `s'. 
  }
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx10_z3_analytic.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx10_z5_analytic.ps}
  \includegraphics[width=\linewidth]{figures/compare_modes_cool_kx10_z7_analytic.ps}
  \caption{Unstable modes with $\khat=10$ and thermal
    relaxation timescales $\beta=0.01$ (black dots), $\beta=0.03$ (red
    crosses), $\beta=0.05$ (green triangles) and $\beta=0.1$ (blue
    crosses); for different vertical domain sizes $\zmax=3H$ (top),
    $\zmax=5H$ (middle, our fiducial setup) and $\zmax=7H$
    (bottom). Lines are computed analytically from
    Eq. \ref{relax_disp} for  $M=0$ (solid, fundamental mode) and
    $M=1,\,2,\,3$ (dotted, dashed, dash-dot, respectively). Along each
    line, $\beta$ increases continously from $0.01$ to $0.1$ from top
    to bottom, and squares mark corresponding $\beta$ values with
    numerical results. 
    \label{compare_modes_cool_kx10} 
  }
\end{figure}

@@

In Fig. \ref{compare_modes_cool_kx30} we plot eigenvalues for unstable
modes with $\khat=30$. For clarity we only plot analytical results for
the fundamental mode, which is again robust against boundary
conditions compared to higher order modes. We find surface modes
appear for $\zmax\geq5H$. However, they are effectively stabilized with
increasing $\beta$. For $\zmax=3H$, the fundamental mode becomes
dominant for $\beta\geq 0.03$, as observed for $\khat=10$. For
$\zmax=5H$ and $\zmax=7H$, we find the emergence modes slightly more
unstable than the fundamental mode even when $\beta=0.1$ (the leftmost
eigenvalues in the lower two panels). However, these modes are clearly associated
with the vertical domain size as they are absent for $\zmax=3H$. 
% Thus, for $\zmax=5H$ and  
% $\zmax=7H$  the fundamental mode is likely dominant for 
% $\beta\geq0.05$

@@

Fig. \ref{compare_modes_cool_kx10} and \ref{compare_modes_cool_kx30} 
show that the fundamental mode is not sensitive to boundary
conditions. The fundamental mode also becomes dominant as $\beta$
approaches $\beta_\mathrm{crit}$. %beta<beta_crit is necessary and sufficient  
Thus, tracking the fundamental mode is a more robust way of 
assessing stability as $\beta$ is increased from zero, rather than
searching for the most unstable mode, which may correspond to modes
that depend on boundary conditions.

>>>>

\subsection{VSI with rapid thermal relaxation}\label{vertiso_pertiso} 
We first calculate the VSI in a disk with rapid thermal relaxation by
setting $\beta=10^{-3}$. (Smaller values yield similar results.) This
is similar to strictly isothermal perturbations in a vertically
isothermal disk considered in \cite{nelson13} and \cite{mcnally14}.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_modes_iso_kx30_analytic.ps}
  \caption{Same as Fig. \ref{compare_modes_iso_kx10} but for $\khat=30$. Examples of surface modes are
    marked by `s'. \label{compare_modes_iso_kx30}
  }
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx10_z3_analytic.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx10_z5_analytic.ps}
  \includegraphics[width=\linewidth]{figures/compare_modes_cool_kx10_z7_analytic.ps}
  \caption{Unstable modes with $\khat=10$ and thermal
    relaxation timescales $\beta=0.01$ (black dots), $\beta=0.03$ (red
    crosses), $\beta=0.05$ (blue triangles) and $\beta=0.1$ (green 
    crosses); for different vertical domain sizes $\zmax=3H$ (top),
    $\zmax=5H$ (middle, the fiducial setup) and $\zmax=7H$
    (bottom). Lines are computed from
    Eq. \ref{relax_disp} for  $M=0$ (solid, fundamental mode) and
    $M=1,\,2,\,3$ (dotted, dashed, dash-dot, respectively). Along each
    line, $\beta$ increases continously from $0.01$ to $0.1$ from top
    to bottom, and squares mark corresponding $\beta$ values with
    numerical results. 
    \label{compare_modes_cool_kx10} 
  }
\end{figure}

@@

In Fig. \ref{compare_modes_cool_kx30} we plot eigenvalues for unstable
modes with $\khat=30$. For clarity we only plot analytical results for
the fundamental mode, which is again insensitive boundary
conditions compared to higher order modes and thus likely more
robust in realistic disks. We find surface modes
appear for $\zmax\geq5H$. However, they are effectively stabilized with
increasing $\beta$. For $\zmax=3H$, the fundamental mode becomes
dominant for $\beta\geq 0.03$, as observed for $\khat=10$. For
$\zmax=5H$ and $\zmax=7H$, we find the emergence modes slightly more
unstable than the fundamental mode even when $\beta=0.1$ (the leftmost
eigenvalues in the lower two panels). However, these modes are clearly associated
with the boundary condition as they are absent for $\zmax=3H$. 
% Thus, for $\zmax=5H$ and  
% $\zmax=7H$  the fundamental mode is likely dominant for 
% $\beta\geq0.05$

@@

Fig. \ref{compare_modes_cool_kx10} and \ref{compare_modes_cool_kx30} 
show that the fundamental mode is not sensitive to boundary
conditions. The fundamental mode also becomes dominant as $\beta$
approaches $\beta_\mathrm{crit}$. %beta<beta_crit is necessary and sufficient  
Thus, tracking the fundamental mode is a more robust assessment of
stability as $\beta$ is increased from zero, rather than 
searching for the most unstable mode, which may correspond to modes
that depend on boundary conditions.
